The evolution of interdisciplinarity and internationalization in scientific journals

Version 1

Abstract

Abstract

There is a widely-held perception that science is becoming increasingly international — both in terms of collaborations involving authors form multiple countries and in terms of publications from authors from a broader group of countries — and more interdisciplinary, i.e., drawing on knowledge and methods from multiple scholarship domains. However, these hypothesized trends have not yet been characterized quantitatively at the level of individual discipline. The widespread availability of scholarly metadata on scientific publications makes it possible to examine how science has evolved over time. Here, we use OpenAlex metadata to examine potential trends in two groups of peer-reviewed research journals: disciplinary journals in biology, chemistry, economics, medicine, physics, and political science, and multidisciplinary journals that publish articles from multiple fields. Supporting existing perceptions, we find an almost universal trend towards increasing internationalization of both sets of journals. Nevertheless, we find disparities: journals in medicine are less international than journals in other disciplines and are not increasing their levels of internationalization, whereas journals in physics appear to be segregating between those that are international and those that are not. We also find that multidisciplinary journals have undergone significant shifts in their disciplinary focuses over the past century, whereas disciplinary journals appear to have largely maintained their degree of interdisciplinarity.

Introduction

Peer-reviewed scientific journals are the backbone of scholarly communication for the worldwide community of scholars [1]. Remarkably, the volume of scholarly journal article publications has been growing exponentially for over a century (See Fig. 1b) [2]. The dominance of journal articles was not always the state of affairs in scientific communication. The longest-running scholarly journal was launched in 1665 (The French Le Journal des Sçavans (or Journal des sąvans) was published for the first time a few months earlier but its publication was interrupted twice [3].) — the Royal Society of London launched the Philosophical Transactions of the Royal Society to collect and distribute important scientific findings and news [4]. Unlike nowadays, in the 18th and 19th centuries monographs, literary essays, public and private presentations, personal letters, pamphlets and full-length books were all significant channels for communicating research findings and established priority [5]. By the turn of the 20th century, printed scientific journals emerged as the primary means of disseminating new scientific discoveries [5, 6]. In this new millennium, digital publications, still within the framework of scholarly journals, have overtaken printed publications [7].

Scholarly journal articles are the largest category of scholarly publications. The number of journal publications increased steadily except during the two World Wars.

Solid lines show fits to the data which have, in some cases, been shifted vertically for clarity. a.g Scholarly works (with relevant metadata) indexed in the OpenAlex database are categorized into seven groups. Most publications are journal articles. b. Number of journal articles has been doubling approximately every 10 years for over a century. While there were significant drops during World War I (1914 to 1918) and World War II (1939 to 1945), the growth of journal article publications quickly recovered. The solid line shows a linear fit to the logarithm of the dependent variable. c. Number of actively publishing journals has been doubling approximately every 14 years for over a century, but the growth rate has not been constant. The two solid lines show the linear fits to the logarithm of the dependent variable to different time periods. d. Number of countries represented in the affiliations of the authors of annually published journal articles. The data is well-described by a logistic curve centered at 1970. Most of the growth in the number of countries occurred between 1949 and 1991.

The scope of scientific enterprise has also evolved over time. In the 380s B.C.E, Aristotle classified all human knowledge into two philosophical categories: speculative and practical [8]. Practical philosophy encompassed topics such as economics, ethics, and politics. Speculative philosophy encompassed aesthetics, mathematics, metaphysics, and physics (also called natural philosophy) [9]. Not until the 19th century did the current partitioning of scholarly endeavors start to take shape [10]. This re-arrangement of human knowledge was accompanied by its exponential growth, prompting increasing specialization of scholars and the emergence of ever more specialized venues for sharing discoveries. Perhaps as a result of this specialization trend, the number of scholarly journals has been doubling approximately every 14 years (Fig. 1c). Much of this growth has occurred within the dominant natural sciences — biology, chemistry, and physics — each now encompassing a multitude of sub-fields [11–13]. Counter-intuitively, increasing specialization has produced a drive for interdisciplinarity, with researchers increasingly exchanging knowledge and methods across domains and bridge disciplinary boundaries [14–17]. Due to this drive, there had been a second trend in opposition to ever more specialized journals: the increasing prominence of multidisciplinary journals (e.g. Nature, Science, and PNAS), which feature research contributions from a multiple disciplines [11,16]. Another factor that has transformed the worldwide scientific enterprise is its increasing internationalization, with scientific collaborations now extending across national boundaries. Whereas France, Germany, the United Kingdom (U.K.), and the United States (U.S.), dominated knowledge production in the 19th and 20th centuries, in recent decades we have witnessed the increasing importance to knowledge creation of Asian nations, such as China, India, and Japan (Fig. S1 in Extended Data: Statistics on journal publications at the country level).

In fact, there has been a significant expansion in the number of countries where scientists are publishing in selective journals. As seen in Fig. 1d), these expansion is well described by a logistic model with a growth rate that peaked in 1970, but extends over the period 1949 to 1991. Plausibly, the widespread focus on increased scientific research was due to the importance of science and technology to the outcomes of the World Wars.

We surmise that these perceived trends towards greater interdisciplinarity and internationalization will be visible and quantifiable using data-driven approaches. We further hypothesize that quantifying these processes will yield a deeper understanding of the observed changes. These insights will guide researchers and policymakers in promoting collaborations among different disciplines and countries, thus encouraging the type of innovative research necessary to address the complex challenges currently faced by humanity.

Data and Methods

We focused our analysis on the March 2022 snapshot of OpenAlex, a large-scale and open scholarly metadata source. When Microsoft ceased updates of the Microsoft Academic Graph (MAG), at the end of 2021 [18], the non-profit organization OurResearch [19] incorporated the entire MAG corpus — excluding patent data — into OpenAlex [20]. Hence, the snapshot of OpenAlex we studied maintains all features and data schemas used by the MAG.

The March 2022 snapshot of OpenAlex contains information on 208,755,206 scholarly works. OpenAlex categorizes scholarly works into seven types: journal articles, conference proceedings, repositories, book chapters, books, theses, and datasets (Fig. 1a). We refined the downloaded data by filtering out entries lacking essential information for our analysis, such as publication year, document type, authorship, affiliation, and field of study information. We then restricted our attention further to scholarly works published after 1900 (due to poor coverage of earlier works), and before 2021 (due to incomplete inclusion of more recent works). This filtering procedure left us with a corpus comprising 56,697,402 publication records, 87% of which are journal articles.

These ~50 million articles were published across 51,062 scholarly journals. Remarkably, Fig. 1b shows that the number of journal articles published annually has been growing exponentially with an astonishing doubling period of a decade. Even the significant impact on publication rates of the two World Wars were quickly overcame, and exponential growth resumed.

As journals comprise the largest share of scholarly publications, we choose to study how two characteristics of journal articles — such as diversity of topics and countries — evolved over the last century. A significant challenge with these records, however, is the frequent unavailability of authorship or affiliation metadata —-a limitation that also affected the MAG (see Fig. S2 for a comparison to other databases). The reasons behind this missing metadata in OpenAlex remain unclear. Due to the difficulty of accurately imputing these missing affiliations, our analysis focuses exclusively on journal article publications that include both authorship and affiliation metadata. Approximately 54% of journal articles in the database have complete authorship and affiliation metadata.

Metadata

Every journal article recorded by OpenAlex is assigned to a unique identifier: ‘PaperId’, and the journal where it was published is also assigned to a unique identifier:’JournalId’ [20]. These internal identifiers also connect to external identifiers, such as the International Standard Serial Number (ISSN) [21], the electronic ISSN (eISSN) [21], and the Research Organization Registry Identifier (ROR ID) [22], enabling us to connect information across multiple external data sources. In particular, we can connect authors’ affiliations to institutions listed in the ROR database [22] and thus identify the country where the institution is located.

We then refered to the Clarivate Analytics’ Journal Citation Reports [23] to obtain the disciplinary classification of the journals in our corpus [24, 25]. We selected for detailed investigation 7 multidisciplinary journals and 16 journals for each of six disciplines across the natural sciences (Chemistry and Physics); life sciences (Biology and Medicine); and social sciences (Economics and Political Science). We selected journals with long publishing history, large number of articles published annually, high impact factors (based on the journal impact factor in 2022), and restricted our attention to journals primarily publishing original research, thus excluding journals focusing on reviews. We then obtained the full publication records of each journal from OpenAlex corpus [20] (see Extended Data: Fields of study and Fig. S3).

Quantification

The interdisciplinarity of a journal is measured by the extent to which its articles draw upon knowledge and methods from diverse disciplines. To quantify this, we make use of the fact that we can calculate the probability distribution pjt(d) of the primary — i.e., level L0 (see Extended Data: Field of Study for details) — disciplines d of the articles published in a given journal j in a given year t. Since a single article may cover more than one discipline, indicating the integration of knowledge and methods from different fields of study, we compute a fractional discipline count for every article, sum across all articles published in the journal in the specified year, and divide by the total number of articles so as to get the probability. We use pjt(d) to define an interdisciplinarity index Id(j, t) by making use of Shannon’s entropy [26], which captures the diversity of disciplinary representation in the journal’s content:

A higher index value indicates greater diversity in the disciplinary focus of a journal, while a lower index value suggests a more concentrated disciplinary focus. Figure S4 demonstrates that higher values of Id are due to more papers drawing from multiple disciplines than to sing discipline papers from multiple disciplines being published in the journal.

Similarly, to quantify the the internationalization of a journal, we make use of the fact that we can calculate the probability distribution pjt(n) of the countries n of affiliation of the authors of the articles published in a given journal j in a given year t. Since a single article may include affiliations from more than one country, we calculate a fractional country count for every article, sum across all articles published in the journal in the specified year, and divide by the total number of articles so as to get the probability. We use pjt(c) to define an internationalization index In(j, t) by making use of Shannon’s entropy [26]:

A higher index value indicates greater diversity in the country affiliations of the authors of the articles published in a journal, while a lower index value suggests a more restricted set of countries. Figure S4 suggests that higher values of In are not necessarily due to more papers authored by researchers from multiple countries but can be due (see case of Chemistry) to papers from authors from a single country but with greater variety of the specific country.

Shannon’s entropy is one of many measures widely used to quantify deviation from random expectation [27,28]. We chose it from among other measures because it has nice properties and its value can be interpreted as the “effective” number of different categories in the data.

Results

Top multidisciplinary journals

To build confidence on our metrics, we start by investigating the interdiciplinarity and internationalization of three of the most highly regarded multidisciplinary journals — Nature, Science, and The Proceedings of the National Academy Sciences of the United States of America (PNAS). All three journals have a long history and publish a large number of articles annually, mostly original research.

Figure 2a shows the temporal evolution of the interdisciplinarity index for these three journals. It is visually apparent that Nature and PNAS, in particular, have experienced significant shifts in the degree of interdisciplinarity of the articles they publish. To better understand the specifics of these changes, we next explore how different disciplines received evolving degrees of attention in each of these journals over time (Figs. 2b-d). Because of their preponderance in the data, we split Biology, Chemistry, and Physics articles into separate groups and pool all articles on other major fields of study into an “Other Fields” group.

Long-standing, high reputation multidisciplinary journals have shifted their disciplinary focus quite dramatically over the last 100 years.

a. Time evolution of the interdisciplinarity indices, Id, of Nature, Science, and PNAS, measured by the entropy of the annual distribution of disciplines in published articles. b-d. Time evolution of the disciplinary partitioning of articles published in PNAS (b), Nature (c), and Science (d). For all three journals, there has been a shift from ‘Other Fields’ to Physics, Chemistry and Biology. In the 1920s, nearly 60% of articles in Nature or Science were in fields other than Physics, Chemistry and Biology. Interestingly, after following different paths, all three journals are now publishing nearly 40% of their articles in fields other than Physics, Chemistry and Biology. In PNAS, there was a noticeable shift towards Biology from 1960 to 1990 and away from Physics and “Other Fields”. After 1990, “Other Fields” grew at the expense of Biology. In Nature, there has been a steady growth of Biology focused articles. For both Nature or Science, the number of Chemistry publications began its ascent in 1940, reaching a pinnacle around 1970, and a slower decline after that.

Upon this breakdown, a clear trend emerges across all three journals. An initial period where a broad range of disciplines are published in those journals, followed by a transition to a greater emphasis on Biology, Chemistry, and Physics. Specifically, in the 1920s, nearly 60% of articles in Nature and Science concerned disciplines other than Biology, Chemistry, or Physics (Figs. 2c,d). By 1990, approximately 90% of articles in PNAS reported on either Biology or Chemistry studies (Fig. 2b). For Nature, the peak concentration was reached in the 1970s, when 60% of articles reported on either Biology or Chemistry studies. Despite distinct historical trajectories, all three journals have evolved in recent years toward a situation where nearly 40% of the articles they publish fall outside of the dominant disciplines, but where Biology studies account for 20–30% of all publications.

Figure 3a shows the time evolution of the internationalization index for Nature, PNAS, and Science. It is visually apparent that PNAS and Science, in particular, have experienced an extraordinary growth in the diversity of countries from which authors of published articles originate. To better understand the specifics of these aggregate trends, we next explore how authors affiliated with institutions located in countries from different regions were increasingly able to publish in these leading multidisciplinary journals (Figs. 3b-d). Because of their preponderance in the data, we group countries into three main regions: Asia, Europe, and North America. We group all affiliations to institutions outside of those regions into “Other Regions.”

Long-standing, high reputation multidisciplinary journals have become more international.

a. Time evolution of the internationalization indices, In, of Nature, Science, and PNAS, measured by the entropy of location of the affiliations of the authors of published articles. The index has increased drastically over the last 100 years for all three journals. b-d. Time evolution of the percentage of articles whose authors are affiliated with institutions from Asia, Europe, or North America in PNAS (b), Nature (c), and Science (d). For all three journals, the greatest growth has occurred for authors with affiliations in institutions located in Asia. Surprisingly, PNAS has now the largest fraction of authors from Asian institutions, even though Nature had a higher level in the 1940s. For both Science and PNAS, the fraction of authors from European institutions has increased dramatically since the 1960s. Around the same time, the fraction of authors from European institutions publishing in Nature decreased dramatically as the fraction of authors from North American institutions grew rapidly.

Upon this breakdown, a clear trend emerges for the two U.S. based journals. Both PNAS and Science started as a publication outlet for only authors with affiliations in North American institutions (Figs. 3b,d). Starting in the 1980s, both experience an increase in the fraction of papers authored by researchers affiliated with European institutions and, to a lesser extent, Asian institutions. Currently, about 60% of papers published in PNAS and Science are authored by researchers affiliated with North American Institutions, but that percentage appears likely to continue to decline.

The story is more complex for Nature (Fig. 3c). Already over a century ago, authors from institutions in Europe and North American were publishing in this U.K.-based journal. Moreover, already by the 1940s, authors affiliated with institutions in Asia started publishing in Nature. Despite this early internationalization effort, Nature now publishes about the same fractions of Asian, European and North American authors as Science, but a smaller fraction of authors affiliated with institutions from Asia than PNAS.

This internationalization shift may be partially attributed to external factors such as the rise of electronic publishing which now offers researchers a multitude of platforms for communicating their research and an easier path to submitting manuscripts to — or reviewing manuscripts for — journals based in other continents. Electronic publishing also enabled a new model — open access [7, 29]. Initial studies suggested that open access publication produced greater research visibility and greater number of citations [30]. Another important factor impacting journal internationalization is the growth of team sizes [31, 32] and of international collaborations as more diverse collaborating teams may prioritize different factors in the choice of publication venue.

Disciplinary and Multidisciplinary journals: Interdisciplinarity

Next, we compare the changes in interdisciplinarity and internationalization for disciplinary journals (see Data and Methods for details). Reassuringly, our analysis shows that over the observed time periods, the interdisciplinarity index (Id) of multidisciplinary journals consistently remains significantly higher than that of disciplinary journals (Fig. 4, Extended Data: Interdisciplinarity dynamics, and Fig. S5). Considering just the most recent indices (as shown in Table S1), we found that the average interdisciplinarity index for multidisciplinary journals is approximately 3 (Science has the highest interdisciplinarity index, Id ≈ 3.5). Moreover, the interdisciplinarity of multidisciplinary journals appears to reach as steady-state after 1990.

Trends in interdisciplinarity across journals in natural sciences, life sciences, social sciences, and multidisciplinary sciences in 1970, 1990, 2010, and 2020.

Most recent interdisci-plinarity index at a specific year, Id, versus the linear trend of the journal’s interdisciplinarity index over the preceding 5 years. Multidisciplinary journals consistently exhibit a substantially higher in-terdisciplinarity index (Id ≈ 3) compared to all disciplinary journals throughout the observed period (Table S1). Disciplinary journals in natural science (Chemistry and Physics) and life science (Biology and Medicine) appear to fall into two groups based on their interdisciplinarity indices: The first group of journals maintains a narrow focus, publishing research that draws on a single discipline (Id ≈ 0). The second group of journals publishes moderately interdisciplinary research (Id ≈ 1.5). This separation between mono-disciplinary and moderately interdisciplinary journals is visually apparent for journals in natural and life sciences. In contrast, the disciplinary journals in Economics and Political Science we studied seem to be moderately interdisciplinary (Id ≈ 1.5).

The evolution of interdisciplinarity is more diverse for disciplinary journals. Physics and Chemistry journals display intriguing changes in interdisciplinarity. A majority of those journals has increase dramatically in interdisciplinarity over time, with Physics journals, in particular, reaching very high degrees of interdisciplinarity. In contrast, a smaller subset of journals have become very mono-disciplinarity. Consistent with this finding, we categorize disciplinary journals into two groups according to their interdisciplinarity index: mono-disciplinary (Id ≈ 0) and moderately interdisciplinary (Id ≈ 1.5).

A segregation between moderately interdisciplinary and mono-disciplinary journals is not as apparent for Biology and Medicine journals. The small values of the 5-year trends show that the inter-disciplinarity of Biology and Medicine journals is also more stable than that of Physics and Chemistry journals.

Economics journals show a trend similar to that of the journals in the natural and life sciences. That is, their interdisciplinarity has increase over time and there is a spread in the levels of interdisci-plinarity around moderate interdisciplinarity. In contrast, Political Science journals have maintained high levels of interdisciplinarity for the entire period studied — a level that was achieved only recently by the most interdisciplinary Physics journals. Perhaps this is not surprisingly as political science is a multifaceted discipline that incorporates insights from fields as diverse as business, economics, law, psychology, and sociology [33].

Thus, despite there being an overall perception of a growing emphasis on interdisciplinary research, the reality is that some disciplinary journals have strengthened their mono-disciplinary focus. The reasons for this are likely complex and multifaceted but may be, at least in part, due to traditional academic structures and publishing norms that favored depth over breadth. Indeed, some of the mono-disciplinary journals in Biology, Chemistry and Physics are highly regarded journals — Molecular Cell (Id ≈ 0.03), Journal of the American Chemical Society (Id ≈ 0.3), and Nature Physics (Id ≈ 0.3). However, we also observe a statistically significant positive correlation between the latest value of Id and its most recent 5-year trend for Biology journals (Table S1) suggesting that Biology journals which are already interdisciplinary are continuing to increase their degree of interdisciplinarity. The emergence of novel areas such as Systems and Synthetic Biology may be responsible for this trend [34].

Disciplinary and Multidisciplinary journals: Internationalization

We next compare the inter-nationalization index of disciplinary and multidisciplinary journals over time (Fig. 5, Extended Data: Internationalization dynamics and Fig. S7). With regards to internationalization, multidisciplinary journals are not the trend setters (Table S2). Instead, disciplinary journals in Physics have the highest In, while disciplinary journals in Biology and Chemistry have internationalization indices comparable to that of multidisciplinary journals. An exception to this pattern are the two mega open-access journals: PLOS One and Scientific Reports, which have internationalization indices as high the the most internationalized Physics journals. The remaining five multidisciplinary journals exhibit interna-tionalization indices comparable to those of Biology and Chemistry journals. This finding is perhaps unsurprising, given the substantial representation of biological and chemical research within these multidisciplinary journals.

Trends in internationalization across journals in natural sciences, life sciences, social sciences, and multidisciplinary sciences in 1970, 1990, 2010, and 2020.

Most recent internationalization index at a specific year, In, versus the linear trend of the journal’s internationalization index over the preceding 5 years. The internationalization of disciplinary journals exhibits a variety of behaviors (see also Table S2). Physics journals have steadily increased their internationalization, which is now higher than for most multidisciplinary journals. comparable to those of multidisciplinary journals during this period. In 2020, the mean In of Physics journals is approximately 4.1. Journals in Biology and Chemistry display similar trends, but with lower mean In, approximately 3.6 for both Chemistry and and Biology journals. In contrast, while a few Medicine and Economics journals have In ≈ 4, most journals in these disciplines and in Political Science have In < 3.

The very high degree of internationalization of Physics journals is also not surprising. Large international initiatives such as the Conseil Européen pour la Recherche Nucléaire (CERN) colliders [35], the Super-Kamiokaze neutrino detection facility [36], the Laser Interferometer Gravitational-Wave Observatory (LIGO) [37] and the Paranal Observatory [38] have been a constant of the Physics research landscape since the 1950s.

Our data also reveals a statistically significant correlation between the latest value of In and its most recent 5-year trend for disciplinary journals in Physics. This correlation is consistent with idea that Physics journals that are already highly internationalized are becoming even more internationalized, while the least internationalized are becoming less internationalized. This trend highlights a growing divide between journals actively pursuing internationalization and those either unwilling or unable to do so.

Similarly to Physics, biological research been revolutionized by global initiatives such as the Human Genome Project (HGP) [39], the International HapMap Project [40], or the 1000 Genomes Project (1KGP) [41]. Remarkably, disciplinary journals in Chemistry have achieved a similarly high degree of internationalization without the benefit of large international collaborations. The explanation here may be the need global spread of the fossil fuels that are so critical to the chemical industry and have promoted development of the discipline across the globe [42].

Among disciplinary journals, journals in Medicine exhibit the lowest level of internationalization (mean In ≈ 1.9). This finding aligns with prior reports that highlight the relatively lower levels of international collaboration in applied and clinical medicine [43]. There are, however, two notable exceptions: The Lancet (In ≈ 4.0) and the European Journal of Clinical Investigation (In ≈ 4.4). Since most Medical journals maintain a modestly low internationalization index, we surmise that medical research adopts a predominantly regional focus and field-specific norms, such as localized health issues, regional regulatory standards, etc.

Surprisingly, some disciplinary journals in Political Science also have low internationalization (mean In ≈ 2.7). Again, this may be due to differences in what type of questions researchers from different countries may want to investigate.

Disciplinary journals: Impact of team size

Since very large collaborations are not unusual in Biology, Medicine or Physics (Table S3), we investigated whether our findings for those two disciplines could have been due to the characteristics of the largest teams. As a control, we repeated the previous analyses for these three disciplines but excluding all publications with 10 or more authors.

Consistent with other studies, we found that there has been a consistent and significant rise in the number of publications authored by teams of 10+ individuals. This increase was already visible in Physics and Medicine by the 1960s. For Biology, the increase becomes noticeable after 1980 (Fig. S8).

Our control analyses confirm the results obtained when considering all teams, thus demonstrating that the trends in interdisciplinarity and internationalization are not explainable by the presence of very large teams in those disciplines (Fig. S9). That is, the observed internationalization of the research published in these top disciplinary journals occurs because the disciplines have created a truly international enterprise.

Discussion

Our findings highlight a transformational shift within the landscape of scientific publishing — the increasingly interconnected nature of scientific research both at the international and interdisciplinary levels. The reasons for this increased interconnectedness are multifaceted. Increased journal competition for the most exciting research, the need for access to scarce or unique resources, the desire to attract a talented workforce. Whatever the reasons, it is critical to recognize the need for such interconnectedness if one aims to tackle the enormous societal challenges that face us, from climate change to increased conflict.

While this shift may bring hope to some, it may be seen as threatening by others. Either because of a misguided focus on competition between nations and the need to retain claims of primacy, or because of wishes to preserve disciplinary purity. Indeed, some current efforts in some countries to “safeguard their national research enterprises” may prove costly to the creation of new knowledge within their own nations. Whatever the motivation, it would be wise for policy makers, research funding organizations, and research institutions to understand and adapt to this shifting landscape in order to effectively promote innovation.

Our study is not without limitations. Perhaps the greatest limitation is that we focus on disciplines where journal articles are the primary scholarly communication channel. This limitation prevents us from being able to generalize our findings to disciplines in the Humanities, which rely on books or book chapters, or in Computer Science and Engineering, which rely on Conference Proceedings [44]. A second limitation of our study is the fact that we only consider 7 multidisciplinary journals and 16 journals for each discipline. The reason for the small number of multidisciplinary journals considered is that there are not that many highly respected journals thus classified. Nonetheless, the journals selected tend to publish large numbers of articles and to have high reputation (see Data and Methods for details). While the requirements we impose limit the number of disciplinary journals that we can consider, we believe that the selected journals provide an accurate reflection of the changes occurring within the disciplines we consider.

eLife Assessment

This important study uses data on over 56 million articles to examine the dynamics of interdisciplinarity and international collaborations in research journals. The data analytics used to quantify disciplinary and national diversity are convincing, and support the claims that journals have become more diverse in both aspects.

Reviewer #1 (Public review):

Summary:

The authors aim to explore how interdisciplinarity and internationalization-two increasingly prominent characteristics of scientific publishing-have evolved over the past century. By constructing entropy-based indices from a large-scale bibliometric dataset (OpenAlex), they examine both long-term trends and recent dynamics in these two dimensions across a selection of leading disciplinary and multidisciplinary journals. Their goal is to identify field-specific patterns and structural shifts that can inform our understanding of how science has become more globally collaborative and intellectually integrated.

Strengths and Weaknesses:

The paper's primary strength lies in its comprehensive temporal scope and use of a rich, openly available dataset covering over 56 million articles. The interdisciplinary and internationalization indices are well-founded and allow meaningful comparisons across fields and time. Moreover, the distinction between disciplinary and multidisciplinary journals adds valuable nuance. However, some methodological choices, such as the use of a 5-year sliding window to compute trend values, are insufficiently justified and under-explained. The paper also does not fully address disparities in data coverage across disciplines and time, which may affect the reliability of historical comparisons. Finally, minor issues in grammar and clarity reduce the overall polish of the manuscript.

Evaluation of Findings:

Overall, the authors have largely succeeded in achieving their stated aims. The findings-such as the sharp rise in internationalization in fields like Physics, and the divergence in interdisciplinarity trends across disciplines-are clearly presented and generally well-supported by the data. The authors effectively demonstrate that scientific journals have not followed a uniform trajectory in terms of structural evolution. However, greater clarity in trend estimation methods and better acknowledgment of dataset limitations would help to further substantiate the conclusions and enhance their generalizability.

Impact and Relevance:

This study makes a timely and meaningful contribution to the fields of scientometrics, sociology of science, and science policy. Its combination of scale, historical depth, and field-level comparison offers a useful framework for understanding changes in scientific publishing practices. The entropy-based indicators are simple yet flexible, and the use of open bibliometric data enhances reproducibility and accessibility for future research. Policymakers, journal editors, and researchers interested in publication dynamics will likely find this work informative, and its methods could be applied or extended to other structural dimensions of scholarly communication.

Reviewer #2 (Public review):

Summary:

This paper uses large-scale publication data to examine the dynamics of interdisciplinarity and international collaborations in research journals. The main finding is that interdisciplinarity and internationalism have been increasing over the past decades, especially in prestigious general science journals.

Strengths:

The paper uses a state-of-the-art large-scale publication database to examine the dynamics of interdisciplinarity and internationalism. The analyses span over a century and in major scientific fields in natural sciences, engineering, and social sciences. The study is well designed and has provided a range of robustness tests to enhance the main findings. The writing is clear and well organized.

Weaknesses:

While the research provides interesting perspectives for the reader to learn about the trends of journal preferences, I have a few points for the authors to consider that might help strengthen their work.

The first thing that comes to mind is the epistemic mechanism of the study. Why should there be a joint discussion combining internationalism and interdisciplinarity? While internationalism is the tendency to form multinational research teams to work on research projects, interdisciplinarity refers to the scope and focus of papers that draw inspiration from multiple fields. These concepts may both fall into the realm of diversity, but it remains unclear if there is any conceptual interplay that underlies the dynamics of their increase in research journals.

It is also unclear why internationalization is increasing. Although the authors have provided a few prominent examples in physics, such as CERN and LIGO, which are complex and expensive experimental facilities that demand collective efforts and investments from the global scientific community, whether some similar concerns or factors drive the growth of internationalism in other fields remains unknown. I can imagine that these concerns do not always apply in many fields, and the authors need to come up with some case studies in diverse fields with some sociological theory to support their empirical findings.

The authors use Shannon entropy as a measure of diversity for both internationalism and interdisciplinarity. However, entropy may fail to account for the uneven correlations between fields, and the range of value chances when the number of categories changes. The science of science and scientometrics community has proposed a range of diversity indicators, such as the Rao-Stirling index and its derivatives. One obvious advantage of the RS index is that it explicitly accounts for the heterogeneous connections between fields, and the value ranges from 0 to 1. Using more state-of-the-art metrics to quantify interdisciplinarity may help strengthen the data analytics.

Author response:

The following is the authors’ response to the original reviews.

Reviewer #1 (Public review):

However, some methodological choices, such as the use of a 5-year sliding window to compute trend values, are insufficiently justified and under-explained. The paper also does not fully address disparities in data coverage across disciplines and time, which may affect the reliability of historical comparisons. Finally, minor issues in grammar and clarity reduce the overall polish of the manuscript.

We thank the reviewer for pointing out the weakness of the manuscript. We addressed these comments in our response to Recommendations A and B. Minor grammar and clarity issues have also been addressed.

Reviewer #2 (Public review):

The first thing that comes to mind is the epistemic mechanism of the study. Why should there be a joint discussion combining internationalism and interdisciplinarity? While internationalism is the tendency to form multinational research teams to work on research projects, interdisciplinarity refers to the scope and focus of papers that draw inspiration from multiple fields. These concepts may both fall into the realm of diversity, but it remains unclear if there is any conceptual interplay that underlies the dynamics of their increase in research journals.

We thank the reviewer for pointing out the lack of clarity in our decision to conduct a joint discussion of interdisciplinarity and internationalization.

It is a well-known fact that team science has increased in importance over time. An important question then is whether teams have only grown in size and frequency or whether they have changed in other aspects. Interdisciplinarity and internationalization are two aspects in which teams could have changed.

We revised the Introduction (Lines 68–70 of the revised manuscript) to address this matter.

It is also unclear why internationalization is increasing. Although the authors have provided a few prominent examples in physics, such as CERN and LAGO, which are complex and expensive experimental facilities that demand collective efforts and investments from the global scientific community, whether some similar concerns or factors drive the growth of internationalism in other fields remains unknown. I can imagine that these concerns do not always apply in many fields, and the authors need to come up with some case studies in diverse fields with some sociological theory to support their empirical findings.

We thank the reviewer for requesting further evidence concerning why our findings may be correct. Physics is an area where the need for extraordinary resources has naturally led to large international collaborative efforts. As we discuss in line 255 of the revised manuscript, this is actually also the case for biology. The Human Genome Project and subsequent projects have also required massive investments, leading to further internationalization.

We believe that the drive toward internationalization for medicine has to do with the need for establishment of robust results that are not specific to a single country or medical system. Additionally, the impact of global epidemics — Acquired immunodeficiency Syndrome (AIDS), Severe Acute Respiratory Syndrome (SARS) — has also increased the needs to involve researchers from around the world.

The case for increased internationalization in the social sciences is, we believe, related to the desire to identify phenomena that extend beyond the Western, educated, industrialized, rich and democratic (WEIRD) societies.

We have expanded the discussion around these points in lines 274–283 of the revised manuscript.

The authors use Shannon entropy as a measure of diversity for both internationalism and interdisciplinarity. However, entropy may fail to account for the uneven correlations between fields, and the range of value chances when the number of categories changes. The science of science and scientometrics community has proposed a range of diversity indicators, such as the RaoStirling index and its derivatives. One obvious advantage of the RS index is that it explicitly accounts for the heterogeneous connections between fields, and the value ranges from 0 to 1. Using more state-of-the-art metrics to quantify interdisciplinarity may help strengthen the data analytics.

We thank the reviewer for pointing the need to provide a deeper discussion of the impact of different metrics on how disciplinary diversity is calculated. We chose Shannon’s entropy because it accounts for both richness (the number of distinct fields) and evenness (the balance of representation across fields). While measures such as the Rao-Stirling index can be very useful when considering disciplines at different levels of aggregation, since to consider only level 0 Field-of-Study (FoS) tags, that problem is not as much a concern for our analysis.

We have added a further clarification in lines 145–151 of the revised manuscript.

Reviewer #1 (Recommendations for the authors)

Ambiguity in the Trend Calculation Methodology in Figure 4 and 5

The manuscript uses a 5-year sliding window to calculate recent trends in interdisciplinarity (Id) and internationalization (In), but the method is not clearly described. Could the authors clarify whether the trend is calculated by (1) performing linear regression on the index values over the past 5 years, (2) using the regression slope as the trend value, and (3) interpreting the sign and magnitude of the slope to indicate increasing, decreasing, or stable trends? Additionally, the rationale for choosing a 5-year window over other durations (e.g., 10 or 15 years) is not discussed. Given that different time windows could yield different insights, a brief justification or sensitivity check would strengthen the methodological transparency.

Thank you for pointing the lack of clarity in our description. In an attempt to increase clarity, we added a specific case study to illustrate the use of 5-year trend in the Supplementary Information: Estimation of tendency of the revised manuscript (Lines 691–704 of the revised manuscript).

Specifically, imagine we want to calculate the trend of the Interdisciplinarity Index for 2010 for Annalen der Physik. We would perform an ordinary least squares linear fit to the 6 data points for the Index in years 2005–2010.

The reason to focus on a 5-year window is two-fold. First, a longer time period would — as suggested by the data on Figure S10 — likely aggregate over multiple trends. Second, a shorter time period would result in too great an uncertainty in the estimation of the trend.

This is the reason why we did not implement a sensitivity analysis. Reasonable time windows that consider the two reasons expressed above would be too narrow to provide a worthwhile analysis.

Lack of Discussion on Temporal Coverage Disparities Across Disciplines

The study spans publications from 1900 to 2021, but the completeness and representativeness of the data-especially in earlier decades-may differ significantly across disciplines. For instance, OpenAlex has limited coverage for publications before the mid-20th century, and disciplines such as Medicine and Political Science may have adopted journal-based publishing at different historical periods compared to Physics or Chemistry. These temporal disparities could bias cross-disciplinary comparisons of long-term trends in interdisciplinarity and internationalization. I recommend that the authors briefly discuss this limitation and, if possible, report when coverage becomes reliable for each discipline. A sensitivity analysis starting from a common baseline year (e.g., 1950 or 1970) could also help assess whether the observed disciplinary differences are driven in part by unequal temporal data availability.

We thank the reviewer for the requesting further clarification on this matter. We completely agree that “completeness and representativeness of the data – especially in earlier decades-may differ significantly across disciplines”. That is exactly the reason why we made the analyses choices described in the manuscript.

Indeed, we consider only three journals for the analysis of the entire 1900–2021 period. Those 3 journals, Nature, PNAS and Science are ones that we know to be well recorded.

When conducting the disciplinary analysis, we focus on the period 1960–2021. While we know that the coverage for the social sciences is less robust until the 1990s, we address this concern by implementing several safeguards:

Manual selection of representative journals in each discipline to ensured that their publications are well represented in OpenAlex.

Decade by decade analysis of interdisciplinarity and internationalization so that changes over time can be identified and potential issues with data coverage are restricted to only some aspects of the analysis.

We also acknowledge the potential coverage disparities in earlier years of the data source (Lines 319-326 of the revised manuscript).

The authors use both interdisciplinarity and multidisciplinarity. While these concepts offer similar definitions of diversity, it may help the reader if there is some explanation to clarify their subtle differences. (Reviewer #2)

It is a well-known fact that team science has increased in importance over time. An important question then is whether teams have only grown in size and frequency or whether they have changed in other aspects. Interdisciplinarity and internationalization are two aspects in which teams could have changed.

We revised the Introduction (Lines 68–70 of the revised manuscript) to address this matter.

Minor Comments

Several sentences

(1) Line 11: The phrase “authors form multiple countries” contains a typographical error. The word “form” should be corrected to “from” so that the sentence reads: “authors from multiple countries.”

tences and phrases throughout the manuscript could be improved for grammatical accuracy, clarity, and stylistic appropriateness:

(2) Line 63: The clause “these expansion is well described by a logistic model” contains a subject-verb agreement error. “These” should be replaced by the singular demonstrative pronoun “this”, resulting in: “This expansion is well described by a logistic model.”

(3) Line 89: The phrase “were quickly overcame” misuses the verb form. “Overcame” is a past tense form and should be replaced with the past participle “overcome” to match the passive construction. Suggested revision: “were quickly overcome.”

(4) Line 106: The verb “refered” is misspelled. It should be corrected to “referred” for proper past tense. The corrected phrase should read: “we referred to...”

(5) Line 127: The phrase “sing discipline papers” contains a typographical error. “Sing” should be “single”, yielding: “single discipline papers.”

(6) Lines 238–239: The sentence “An exception to this pattern are the two mega open-access journals: PLOS One and Scientific Reports, which have internationalization indices as high the the most internationalized Physics journals.” contains multiple grammatical issues.

First, the subject “An exception” is singular, but the verb “are” is plural; this results in a subject-verb agreement error.

Second, the phrase “the the” includes a typographical repetition.

Third, the comparative construction is incomplete; “as high the the...” is ungrammatical and should use “as high as.”

Suggested revision: “An exception to this pattern is the pair of mega open-access journals— PLOS One and Scientific Reports—which have internationalization indices as high as those of the most internationalized Physics journals.”

(7) Line 254: The sentence “biological research been revolutionized...” lacks an auxiliary verb. To be grammatically correct, it should read: “biological research has been revolutionized...”

(8) Line 258: The phrase “need global spread of...” is syntactically awkward. Depending on the intended meaning, it could be revised to either “the global spread of...” or “the global need for the spread of...” for clarity.

(9) Figure S2 Caption: The term “Microsofe Academic Graph” is a typographical error and should be corrected to “Microsoft Academic Graph.”

(10) Reference [40]: The link “ttps://doi.org/10.1038/nature02168” is missing the “h” in “https.” The corrected version is: “https://doi.org/10.1038/nature02168.”

We appreciate your comments on the grammar and clarity of the manuscript. We have thoroughly reviewed and corrected these issues to improve the overall clarity of the text.

Line 11: We changed the typo “form” to “from”.

Line 63: We changed the sentence to “There has been a significant expansion in the number of countries where scientists are publishing in selective journals”.

Line 89 (Line 93 of the revised manuscript): We revised the sentence as suggested, and the revised sentence becomes “Even the significant impacts on publication rates of the two World Wars were quickly overcome, and exponential growth resumed. ”

Line 106 (Line 110 of the revised manuscript): We changed the typo “refered” to “referred”.

Line 127 (Line 131 of the revised manuscript): We changed the typo “Sing” to “single”.

Lines 238-239 (Lines 245-247 of the revised manuscript): We thank the issues pointed out by the reviewer, and we took the reviewer’s suggested version and changed the original sentence to “An exception to this pattern is the pair of mega open-access journals — PLOS One and Scientific Reports — which have internationalization indices as high as those of the most internationalized Physics journals”.

Line 254 (Line 262 of the revised manuscript): We added the auxiliary verb to the sentence, and the sentence now becomes “biological research has been revolutionized”

Line 258 (Line 266 of the revised manuscript): We changed the phrase to “the global need for the spread of”.

Figure S2 Caption: We corrected the typo of “Microsoft Academic Graph”.

Reference [40]: We corrected the URL of the reference.

Reviewer #2 (Recommendations for author):

Some typos:

(1) Page 2: On page 2, “contributions from a multiple disciplines” and ”these expansion is well described”.

(2) Page 4: “World Wars were quickly overcame”.

(3) Page 5: “to quantify the the internationalization of a journal”.

(4) Page 10: “indices as high the the most internationalized Physics journals”

(5) Page 10: The sentence “indices as high the the most internationalized Physics journals” contains multiple issues. The phrase “the the” is a typographical error, and the comparative construction is incomplete. It should be revised to: “indices as high as those of the most internationalized Physics journals.”

We revised those typographical errors on page 2, 4, 5, and 10 pointed out by the reviewer. We truly thank the reviewer’s critical examination on the syntax of the manuscript.

Page 2: We removed “a” so now the sentence reads: “contributions from multiple disciplines.”

Page 2: We changed the sentence to “There has been a significant expansion in the number of countries where scientists are publishing in selective journals”.

Page 4: We replaced “overcame” with the past participle “overcome” , resulting in: “World Wars were quickly overcome.”

Page 5: The phrase “to quantify the the internationalization of a journal” contains a typographical repetition. We changed it to: “to quantify the internationalization of a journal.”

Page 10: For the sentence “indices as high the the most internationalized Physics journals”, we removed duplicated “the” as a typographical error. We revised the sentence into: “indices as high as those of the most internationalized Physics journals.”

Version 2

Abstract

There is a widely-held perception that science is becoming increasingly international — both in terms of collaborations involving authors from multiple countries and in terms of publications from authors from a broader group of countries — and more interdisciplinary, i.e., drawing on knowledge and methods from multiple scholarship domains. However, these hypothesized trends have not yet been characterized quantitatively at the level of individual discipline. The widespread availability of scholarly metadata on scientific publications makes it possible to examine how science has evolved over time. Here, we use OpenAlex metadata to examine potential trends in two groups of peer-reviewed research journals: disciplinary journals in biology, chemistry, economics, medicine, physics, and political science, and multidisciplinary journals that publish articles from multiple fields. Supporting existing perceptions, we find an almost universal trend towards increasing internationalization of both sets of journals. Nevertheless, we find disparities: journals in medicine are less international than journals in other disciplines and are not increasing their levels of internationalization, whereas journals in physics appear to be segregating between those that are international and those that are not. We also find that multidisciplinary journals have undergone significant shifts in their disciplinary focuses over the past century, whereas disciplinary journals appear to have largely maintained their degree of interdisciplinarity.

Introduction

Peer-reviewed scientific journals are the backbone of scholarly communication for the worldwide community of scholars [1]. Remarkably, the volume of publications in scholarly journals has been growing exponentially for over a century (See Fig. 1b) [2]. The dominance of journal articles was not always the state of affairs in scientific communication. The longest-running scholarly journal was launched in 1665 (The French Le Journal des SÇavans [or Journal des sąvans] was published for the first time a few months earlier but its publication was interrupted twice [3].) — the Royal Society of London launched the Philosophical Transactions of the Royal Society to collect and distribute important scientific findings and news [4]. Unlike nowadays, in the 18th and 19th centuries monographs, literary essays, public and private presentations, personal letters, pamphlets and full-length books were all significant channels for communicating research findings and established priority [5]. By the turn of the 20th century, printed scientific journals emerged as the primary means of disseminating new scientific discoveries [5, 6]. In this new millennium, digital publications, still within the framework of scholarly journals, have overtaken printed publications [7].

Scholarly journal articles are the largest category of scholarly publications.

The number of journal publications increased steadily except during the two World Wars. Solid lines show the fits to the data which have, in some cases, been shifted vertically for clarity. a. Scholarly works (with relevant metadata) indexed in the OpenAlex database are categorized into seven groups. Most publications are journal articles. b. Number of journal articles has been doubling approximately every 10 years for over a century. While there were significant drops during World War I (1914 to 1918) and World War II (1939 to 1945), the growth of journal article publications quickly recovered. The solid line shows a linear fit to the logarithm of the dependent variable. c. Number of actively publishing journals has been doubling approximately every 14 years for over a century, but the growth rate has not been constant. The two solid lines show the linear fits to the logarithm of the dependent variable to different time periods. d. Number of countries represented in the affiliations of the authors of annually published journal articles. The data is well-described by a logistic curve centered at 1970. Most of the growth in the number of countries occurred between 1949 and 1991.

The scope of scientific enterprise has also evolved over time. In the 380s B.C.E, Aristotle classified all human knowledge into two philosophical categories: speculative and practical [8]. Practical philosophy encompassed topics such as economics, ethics, and politics. Speculative philosophy encompassed aesthetics, mathematics, metaphysics, and physics (also called natural philosophy) [9]. Not until the 19th century did the current partitioning of scholarly endeavors start to take shape [10].

This re-arrangement of human knowledge was accompanied by its exponential growth, prompting increasing specialization of scholars and the emergence of ever more specialized venues for sharing discoveries. Perhaps as a result of this specialization trend, the number of scholarly journals has been doubling approximately every 14 years (Fig. 1c). Much of this growth has occurred within the dominant natural sciences — biology, chemistry, and physics — each now encompassing a multitude of subfields [11–13]. Counter-intuitively, increasing specialization has produced a drive for interdisciplinarity, with researchers increasingly collaborating to exchange knowledge and methods across domains and bridge disciplinary boundaries [14–17]. Due to this drive, there had been a second trend in opposition to ever more specialized journals: the increasing prominence of multidisciplinary journals (e.g. Nature, Science, and PNAS), which feature research contributions from multiple disci-plines [11, 16].

Another factor that has transformed the worldwide scientific enterprise is its increasing internationalization, with scientific collaborations now extending across national boundaries. Whereas France, Germany, the United Kingdom (U.K.), and the United States (U.S.), dominated knowledge production in the 19th and 20th centuries, in recent decades we have witnessed the increasing importance to knowledge creation of Asian nations, such as China, India, and Japan (Fig. S1 in Extended Data: Statistics on journal publications at the country level).

In fact, there has been a significant expansion in the number of countries where scientists are publishing in selective journals. As seen in Fig. 1d, this expansion is well described by a logistic model with a growth rate that peaked in 1970, but extends over the period 1949 to 1991. Plausibly, the widespread focus on increased scientific research was due to the importance of science and technology to the outcomes of the World Wars.

Recent literature has highlighted the importance of team science in the production of knowledge, especially impactful knowledge, in recent decades. A less well-understood aspect in the growth of team science is the impact of interdisciplinarity and internationalization. A driving thrust of this study is the assumption that perceived trends towards greater interdisciplinarity and internationalization will be visible and quantifiable using data-driven approaches. We further hypothesize that quantifying these processes will yield a deeper understanding of the observed changes. These insights will guide researchers and policymakers in promoting collaborations among different disciplines and countries, thus encouraging the type of innovative research necessary to address the complex challenges currently faced by humanity.

Data and Methods

We focused our analysis on the March 2022 snapshot of OpenAlex, a large-scale and open scholarly metadata source. When Microsoft ceased updates of the Microsoft Academic Graph (MAG), at the end of 2021 [18], the non-profit organization OurResearch [19] incorporated the entire MAG corpus — excluding patent data — into OpenAlex [20]. Hence, the snapshot of OpenAlex we studied maintains all features and data schemas used by the MAG.

The March 2022 snapshot of OpenAlex contains information on 208,755,206 scholarly works. OpenAlex categorizes scholarly works into seven types: journal articles, conference proceedings, repositories, book chapters, books, theses, and datasets (Fig. 1a). We refined the downloaded data by filtering out entries lacking essential information for our analysis, such as publication year, document type, authorship, affiliation, and field of study information. We then restricted our attention further to scholarly works published after 1900 (due to poor coverage of earlier works), and before 2021 (due to incomplete inclusion of more recent works). This filtering procedure left us with a corpus comprising 56,697,402 publication records, 87% of which are journal articles.

These ∼50 million articles were published across 51,062 scholarly journals. Remarkably, Fig. 1b shows that the number of journal articles published annually has been growing exponentially with an astonishing doubling period of a decade. Even the significant impacts on publication rates of the two World Wars were quickly overcome, and exponential growth resumed.

As journals comprise the largest share of scholarly publications, we choose to study how two characteristics of journal articles — such as diversity of topics and countries — evolved over the last century. A significant challenge with these records, however, is the frequent unavailability of authorship or affiliation metadata —a limitation that also affected the MAG (see Fig. S2 for a comparison to other databases). The reasons behind this missing metadata in OpenAlex remain unclear. Due to the difficulty of accurately imputing these missing affiliations, our analysis focuses exclusively on journal article publications that include both authorship and affiliation metadata. Approximately 54% of journal articles in the database have complete authorship and affiliation metadata.

Metadata

Every journal article recorded by OpenAlex is assigned to a unique identifier: ‘PaperId’, and the journal where it was published is also assigned to a unique identifier:’JournalId’ [20]. These internal identifiers also connect to external identifiers, such as the International Standard Serial Number (ISSN) [21], the electronic ISSN (eISSN) [21], and the Research Organization Registry Identifier (ROR ID) [22], enabling us to connect information across multiple external data sources. In particular, we can connect authors’ affiliations to institutions listed in the ROR database [22] and thus identify the country where the institution is located.

We then referred to the Clarivate Analytics’ Journal Citation Reports [23] to obtain the disciplinary classification of the journals in our corpus [24, 25]. We selected for detailed investigation 7 multidisciplinary journals and 16 journals for each of six disciplines across the natural sciences (Chemistry and Physics); life sciences (Biology and Medicine); and social sciences (Economics and Political Science). We selected journals with long publishing history, large number of articles published annually, high impact factors (based on the journal impact factor in 2022), and restricted our attention to journals primarily publishing original research, thus excluding journals focusing on reviews. We then obtained the full publication records of each journal from OpenAlex corpus [20] (see Extended Data: Fields of study and Fig. S3).

Quantification

The interdisciplinarity of a journal is measured by the extent to which its articles draw upon knowledge and methods from diverse disciplines. To quantify this, we make use of the fact that we can calculate the probability distribution pjt(d) of the primary — i.e., level L0 (see Extended Data: Field of Study for details) — disciplines d of the articles published in a given journal j in a given year t. Since a single article may cover more than one discipline, indicating the integration of knowledge and methods from different fields of study, we compute a fractional discipline count for every article, sum across all articles published in the journal in the specified year, and divide by the total number of articles so as to get the probability. We use pjt(d) to define an interdisciplinarity index Id(j, t) by making use of Shannon’s entropy [26], which captures the diversity of disciplinary representation in the journal’s content:

A higher index value indicates greater diversity in the disciplinary focus of a journal, while a lower index value suggests a more concentrated disciplinary focus. Figure S4 demonstrates that higher values of Id are due to more papers drawing from multiple disciplines than to single discipline papers from multiple disciplines being published in the journal.

Similarly, to quantify the internationalization of a journal, we make use of the fact that we can calculate the probability distribution pjt(n) of the countries n of affiliation of the authors of the articles published in a given journal j in a given year t. Since a single article may include affiliations from more than one country, we calculate a fractional country count for every article, sum across all articles published in the journal in the specified year, and divide by the total number of articles so as to get the probability. We use pjt(c) to define an internationalization index In(j,t) by making use of Shannon’s entropy [26]:

A higher index value indicates greater diversity in the country affiliations of the authors of the articles published in a journal, while a lower index value suggests a more restricted set of countries. Figure S4 suggests that higher values of In are not necessarily due to more papers authored by researchers from multiple countries but can be due (see case of Chemistry) to papers from authors from a single country but with greater variety of the specific country.

Shannon’s entropy is one of many measures widely used to quantify deviation from random expectation [27, 28]. Indeed, many researchers in scientometrics and science of science have focused on so-called Rao-Stirling indices [29], which can accounts for heterogeneous distances between disciplines and sub-disciplines. Because we are focusing on the highest level in the hierarchy for Fieldof-Study, this is not as much a concern for us. Additionally, the raising 2 to the index values provides the an “effective” number of different categories in the data, something that helps with gaining insight from the results.

Results

Top multidisciplinary journals

To build confidence on our metrics, we start by investigating the interdiciplinarity and internationalization of three of the most highly regarded multidisciplinary journals — Nature, Science, and The Proceedings of the National Academy Sciences of the United States of America (PNAS). All three journals have a long history and publish a large number of articles annually, mostly original research.

Figure 2a shows the temporal evolution of the interdisciplinarity index for these three journals. It is visually apparent that Nature and PNAS, in particular, have experienced significant shifts in the degree of interdisciplinarity of the articles they publish. To better understand the specifics of these changes, we next explore how different disciplines received evolving degrees of attention in each of these journals over time (Figs. 2b-d). Because of their preponderance in the data, we split Biology, Chemistry, and Physics articles into separate groups and pool all articles on other major fields of study into an “Other Fields” group.

Long-standing, high reputation multidisciplinary journals have shifted their disciplinary focus quite dramatically over the last 100 years.

a. Time evolution of the interdisciplinarity indices, Id, of Nature, Science, and PNAS, measured by the entropy of the annual distribution of disciplines in published articles. b-d. Time evolution of the disciplinary partitioning of articles published in PNAS (b), Nature (c), and Science (d). For all three journals, there has been a shift from ‘Other Fields’ to Physics, Chemistry and Biology. In the 1920s, nearly 60% of articles in Nature or Science were in fields other than Physics, Chemistry and Biology. Interestingly, after following different paths, all three journals are now publishing nearly 40% of their articles in fields other than Physics, Chemistry and Biology. In PNAS, there was a noticeable shift towards Biology from 1960 to 1990 and away from Physics and “Other Fields”. After 1990, “Other Fields” grew at the expense of Biology. In Nature, there has been a steady growth of Biology focused articles. For both Nature or Science, the number of Chemistry publications began its ascent in 1940, reaching a pinnacle around 1970, and a slower decline after that.

Upon this breakdown, a clear trend emerges across all three journals. An initial period where a broad range of disciplines are published in those journals, followed by a transition to a greater emphasis on Biology, Chemistry, and Physics. Specifically, in the 1920s, nearly 60% of articles in Nature and Science concerned disciplines other than Biology, Chemistry, or Physics (Figs. 2c,d). By 1990, approximately 90% of articles in PNAS reported on either Biology or Chemistry studies (Fig. 2b). For Nature, the peak concentration was reached in the 1970s, when 60% of articles reported on either Biology or Chemistry studies. Despite distinct historical trajectories, all three journals have evolved in recent years toward a situation where nearly 40% of the articles they publish fall outside of the dominant disciplines, but where Biology studies account for 20–30% of all publications.

Figure 3a shows the time evolution of the internationalization index for Nature, PNAS, and Science. It is visually apparent that PNAS and Science, in particular, have experienced an extraordinary growth in the diversity of countries from which authors of published articles originate. To better understand the specifics of these aggregate trends, we next explore how authors affiliated with in-stitutions located in countries from different regions were increasingly able to publish in these leading multidisciplinary journals (Figs. 3b-d). Because of their preponderance in the data, we group countries into three main regions: Asia, Europe, and North America. We group all affiliations to institutions outside of those regions into “Other Regions.”

Long-standing, high reputation multidisciplinary journals have become more international.

a. Time evolution of the internationalization indices, In, of Nature, Science, and PNAS, measured by the entropy of location of the affiliations of the authors of published articles. The index has increased drastically over the last 100 years for all three journals. b-d. Time evolution of the percentage of articles whose authors are affiliated with institutions from Asia, Europe, or North America in PNAS (b), Nature (c), and Science (d). For all three journals, the greatest growth has occurred for authors with affiliations in institutions located in Asia. Surprisingly, PNAS has now the largest fraction of authors from Asian institutions, even though Nature had a higher level in the 1940s. For both Science and PNAS, the fraction of authors from European institutions has increased dramatically since the 1960s. Around the same time, the fraction of authors from European institutions publishing in Nature decreased dramatically as the fraction of authors from North American institutions grew rapidly.

Upon this breakdown, a clear trend emerges for the two U.S. based journals. Both PNAS and Science started as a publication outlet for only authors with affiliations in North American institutions (Figs. 3b,d). Starting in the 1980s, both experience an increase in the fraction of papers authored by researchers affiliated with European institutions and, to a lesser extent, Asian institutions. Currently, about 60% of papers published in PNAS and Science are authored by researchers affiliated with North American Institutions, but that percentage appears likely to continue to decline.

The story is more complex for Nature (Fig. 3c). Already over a century ago, authors from institutions in Europe and North American were publishing in this U.K.-based journal. Moreover, already by the 1940s, authors affiliated with institutions in Asia started publishing in Nature. Despite this early internationalization effort, Nature now publishes about the same fractions of Asian, European and North American authors as Science, but a smaller fraction of authors affiliated with institutions from Asia than PNAS.

This internationalization shift may be partially attributed to external factors such as the rise of electronic publishing which now offers researchers a multitude of platforms for communicating their research and an easier path to submitting manuscripts to — or reviewing manuscripts for — journals based in other continents. Electronic publishing also enabled a new model — open access [7, 30]. Initial studies suggested that open access publication produced greater research visibility and greater number of citations [31]. Another important factor impacting journal internationalization is the growth of team sizes [32, 33] and of international collaborations as more diverse collaborating teams may prioritize different factors in the choice of publication venue.

Disciplinary and Multidisciplinary journals: Interdisciplinarity

Next, we compare the changes in interdisciplinarity and internationalization for disciplinary journals (see Data and Methods for details). Reassuringly, our analysis shows that over the observed time periods, the interdisciplinarity index (Id) of multidisciplinary journals consistently remains significantly higher than that of disciplinary journals (Fig. 4, Extended Data: Interdisciplinarity dynamics, and Fig. S5). Considering just the most recent indices (as shown in Table S1), we found that the average interdisciplinarity index for multidisciplinary journals is approximately 3 (Science has the highest interdisciplinarity index, Id ≈ 3.5). Moreover, the interdisciplinarity of multidisciplinary journals appears to reach as steady-state after 1990.

Trends in interdisciplinarity across journals in natural sciences, life sciences, social sciences, and multidisciplinary sciences in 1970, 1990, 2010, and 2020.

Most recent interdisciplinarity index at a specific year, Id, versus the linear trend of the journal’s interdisciplinarity index over the preceding 5 years. Multidisciplinary journals consistently exhibit a substantially higher interdisciplinarity index (Id ≈ 3) compared to all disciplinary journals throughout the observed period (Table S1). Disciplinary journals in natural science (Chemistry and Physics) and life science (Biology and Medicine) appear to fall into two groups based on their interdisciplinarity indices: The first group of journals maintains a narrow focus, publishing research that draws on a single discipline (Id ≈ 0). The second group of journals publishes moderately interdisciplinary research (Id ≈ 1.5). This separation between mono-disciplinary and moderately interdisciplinary journals is visually apparent for journals in natural and life sciences. In contrast, the disciplinary journals in Economics and Political Science we studied seem to be moderately interdisciplinary (Id ≈ 1.5).

The evolution of interdisciplinarity is more diverse for disciplinary journals. Physics and Chemistry journals display intriguing changes in interdisciplinarity. A majority of those journals has increase dramatically in interdisciplinarity over time, with Physics journals, in particular, reaching very high degrees of interdisciplinarity. In contrast, a smaller subset of journals have become very monodisciplinarity. Consistent with this finding, we categorize disciplinary journals into two groups according to their interdisciplinarity index: mono-disciplinary (Id≈ 0) and moderately interdisciplinary (Id ≈ 1.5).

A segregation between moderately interdisciplinary and mono-disciplinary journals is not as apparent for Biology and Medicine journals. The small values of the 5-year trends show that the interdisciplinarity of Biology and Medicine journals is also more stable than that of Physics and Chemistry journals.

Economics journals show a trend similar to that of the journals in the natural and life sciences. That is, their interdisciplinarity has increase over time and there is a spread in the levels of interdisci-plinarity around moderate interdisciplinarity. In contrast, Political Science journals have maintained high levels of interdisciplinarity for the entire period studied — a level that was achieved only recently by the most interdisciplinary Physics journals. Perhaps this is not surprisingly as political science is a multifaceted discipline that incorporates insights from fields as diverse as business, economics, law, psychology, and sociology [34].

Thus, despite there being an overall perception of a growing emphasis on interdisciplinary research, the reality is that some disciplinary journals have strengthened their mono-disciplinary focus. The reasons for this are likely complex and multifaceted but may be, at least in part, due to traditional academic structures and publishing norms that favored depth over breadth. Indeed, some of the mono-disciplinary journals in Biology, Chemistry and Physics are highly regarded journals — Molecular Cell (Id ≈ 0.03), Journal of the American Chemical Society (Id ≈ 0.3), and Nature Physics (Id ≈ 0.3). However, we also observe a statistically significant positive correlation between the latest value of Id and its most recent 5-year trend for Biology journals (Table S1) suggesting that Biology journals which are already interdisciplinary are continuing to increase their degree of interdisciplinarity. The emergence of novel areas such as Systems and Synthetic Biology may be responsible for this trend [35].

Disciplinary and Multidisciplinary journals: Internationalization

We next compare the internationalization index of disciplinary and multidisciplinary journals over time (Fig. 5, Extended Data: Internationalization dynamics and Fig. S7). With regards to internationalization, multidisciplinary journals are not the trend setters (Table S2). Instead, disciplinary journals in Physics have the highest In, while disciplinary journals in Biology and Chemistry have internationalization indices comparable to that of multidisciplinary journals. An exception to this pattern is the pair of mega open-access journals — PLOS One and Scientific Reports — which have internationalization indices as high as those of the most internationalized Physics journals. The remaining five multidisciplinary journals exhibit internationalization indices comparable to those of Biology and Chemistry journals. This finding is perhaps unsurprising, given the substantial representation of biological and chemical research within these multidisciplinary journals.

Trends in internationalization across journals in natural sciences, life sciences, social sciences, and multidisciplinary sciences in 1970, 1990, 2010, and 2020.

Most recent internationalization index at a specific year, In, versus the linear trend of the journal’s internationalization index over the preceding 5 years. The internationalization of disciplinary journals exhibits a variety of behaviors (see also Table S2). Physics journals have steadily increased their internationalization, which is now higher than for most multidisciplinary journals. comparable to those of multidisciplinary journals during this period. In 2020, the mean In of Physics journals is approximately 4.1. Journals in Biology and Chemistry display similar trends, but with lower mean In, approximately 3.6 for both Chemistry and and Biology journals. In contrast, while a few Medicine and Economics journals have In ≈ 4, most journals in these disciplines and in Political Science have In < 3.

The very high degree of internationalization of Physics journals is also not surprising. Large international initiatives such as the Conseil Européen pour la Recherche Nucléaire (CERN) colliders [36], the Super-Kamiokaze neutrino detection facility [37], the Laser Interferometer Gravitational-Wave Observatory (LIGO) [38] and the Paranal Observatory [39] have been a constant of the Physics research landscape since the 1950s.

Our data also reveals a statistically significant correlation between the latest value of In and its most recent 5-year trend for disciplinary journals in Physics. This correlation is consistent with idea that Physics journals that are already highly internationalized are becoming even more internationalized, while the least internationalized are becoming less internationalized. This trend highlights a growing divide between journals actively pursuing internationalization and those either unwilling or unable to do so.

Similarly to Physics, biological research has been revolutionized by global initiatives such as the Human Genome Project (HGP) [40], the International HapMap Project [41], or the 1000 Genomes Project (1KGP) [42]. Remarkably, disciplinary journals in Chemistry have achieved a similarly high degree of internationalization without the benefit of large international collaborations. The explanation here may be the global need for the spread of the fossil fuels that are so critical to the chemical industry and have promoted development of the discipline across the globe [43].

Among disciplinary journals, journals in Medicine exhibit the lowest level of internationalization (mean In ≈ 1.9). This finding aligns with prior reports that highlight the relatively lower levels of international collaboration in applied and clinical medicine [44]. There are, however, two notable exceptions: The Lancet (In ≈ 4.0) and the European Journal of Clinical Investigation (In ≈ 4.4). Since most Medical journals maintain a modestly low internationalization index, we surmise that medical research adopts a predominantly regional focus and field-specific norms, such as localized health issues, regional regulatory standards, etc. Nonetheless, it is likely that global epidemics — Acquired immunodeficiency syndrome (AIDS), Severe acute respiratory syndrome (SARS), Middle East respiratory syndrome (MERS), or Coronavirus disease 2019 (COVID-19) — will increase the pressure to internationalize medical research.

The need for an increase in the internationalization of social sciences research may be less obvious. Indeed, even some Political Science have low internationalization (mean In ≈ 2.7). This may be due to differences in what type of questions researchers from different countries may want to investigate. However, there has been a recent interest by research in the Western world in investigating phenomena that extend beyond what is expected from Western, educated, industrialized, rich and democratic (WEIRD) societies.

Disciplinary journals: Impact of team size

Since very large collaborations are not unusual in Biology, Medicine or Physics (Table S3), we investigated whether our findings for those two disciplines could have been due to the characteristics of the largest teams. As a control, we repeated the previous analyses for these three disciplines but excluding all publications with 10 or more authors.

Consistent with other studies, we found that there has been a consistent and significant rise in the number of publications authored by teams of 10+ individuals. This increase was already visible in Physics and Medicine by the 1960s. For Biology, the increase becomes noticeable after 1980 (Fig. S8).

Our control analyses confirm the results obtained when considering all teams, thus demonstrating that the trends in interdisciplinarity and internationalization are not explainable by the presence of very large teams in those disciplines (Fig. S9). That is, the observed internationalization of the research published in these top disciplinary journals occurs because the disciplines have created a truly international enterprise.

Discussion

Our findings highlight a transformational shift within the landscape of scientific publishing — the increasingly interconnected nature of scientific research both at the international and interdisciplinary levels. The reasons for this increased interconnectedness are multifaceted. Increased journal competition for the most exciting research, the need for access to scarce or unique resources, the desire to attract a talented workforce. Whatever the reasons, it is critical to recognize the need for such interconnectedness if one aims to tackle the enormous societal challenges that face us, from climate change to increased conflict.

While this shift may bring hope to some, it may be seen as threatening by others. Either because of a misguided focus on competition between nations and the need to retain claims of primacy, or because of wishes to preserve disciplinary purity. Indeed, some current efforts in some countries to “safeguard their national research enterprises” may prove costly to the creation of new knowledge within their own nations. Whatever the motivation, it would be wise for policy makers, research funding organizations, and research institutions to understand and adapt to this shifting landscape in order to effectively promote innovation.

Our study is not without limitations. Perhaps the greatest limitation is that we focus on disciplines where journal articles are the primary scholarly communication channel. This limitation prevents us from being able to generalize our findings to disciplines in the Humanities, which rely on books or book chapters, or in Computer Science and Engineering, which rely on conference proceedings [45]. A second limitation of our study is the fact that we only consider 7 multidisciplinary journals and 16 journals for each discipline. The reason for the small number of multidisciplinary journals considered is that there are not that many highly respected journals thus classified.

Another limitation we must discuss relates to the completeness and representativeness of the data across disciplines. This concern was foremost in our minds when deciding on the time period for analysis. Indeed, the reason why we consider the period 1900–2021 for only the the three major interdisciplinary journals — Nature, PNAS, Science — has to do precisely with the fact that those journals are the most likely to be accurately covered. When considerring disciplinary journals we restrict our attention to the period 1960–2021. Even then, for some journals and disciplines covered can be spotty prior to 2000. For this reason, we perform analyses that focus on manually selected journals and for specific time points (1970, 1990, 2010, 2020).

Moreover, the selected journals are among the most reputable in their fields and publish large volumes of articles (see Data and Methods for details). While the requirements we impose limit the number of disciplinary journals that we can consider, we believe that the selected journals provide an accurate reflection of the changes occurring within the disciplines we consider.